---
title: "R Notebook - Doublet Detection with scDblFinder"
author: Srivalli Kolla
output: html_notebook
---

# SETUP AND LOAD LIBRARIES

```{r}
library(Seurat)
library(SingleCellExperiment)
library(scDblFinder)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# LOAD FILTERED AND NORMALIZED DATA

```{r}
INPUT_OBJECT_PATH <- "data/processed/combined_filtered_normalized.rds"
SAVE_OBJECT_PATH <- "data/processed/combined_singlets_filtered.rds"

if (!file.exists(INPUT_OBJECT_PATH)) {
    stop("ERROR: Filtered and normalized object not found. Run 02_filtering_normalization.Rmd first!")
}
combined_filtered <- readRDS(INPUT_OBJECT_PATH)

print(paste("Loaded filtered object with", ncol(combined_filtered), "nuclei."))
```

# PREPARE DATA FOR DOUBLET FINDING

Run initial dimensionality reduction and clustering on the filtered data. This is necessary because scDblFinder uses the proximity of cells in reduced space to model doublets.

```{r}
# Use the SCT assay
DefaultAssay(combined_filtered) <- "SCT"

# Set PCs based on typical expectations for scRNA-seq
PC_TO_USE <- 30 

# Run PCA, Neighbors, and initial Clustering (for context and plotting)
combined_filtered <- RunPCA(combined_filtered, verbose = FALSE, npcs = 50)
combined_filtered <- FindNeighbors(combined_filtered, dims = 1:PC_TO_USE)
combined_filtered <- FindClusters(combined_filtered, resolution = 0.5) 
combined_filtered <- RunUMAP(combined_filtered, dims = 1:PC_TO_USE, verbose = FALSE)

# Convert Seurat to SingleCellExperiment (required by scDblFinder)
sce <- as.SingleCellExperiment(combined_filtered, assay = 'SCT')
```

# RUN scDblFinder

```{r}
# Run scDblFinder (batch-aware using Sample_ID)
sce <- scDblFinder(sce, samples = 'Sample_ID', verbose = FALSE)

# Add doublet results back to the Seurat object metadata
combined_filtered$scDblFinder_class <- colData(sce)$scDblFinder.class
combined_filtered$scDblFinder_score <- colData(sce)$scDblFinder.score

print("Doublet marking complete. Annotations added to metadata.")
```

# FILTER OUT DOUBLETS

```{r}
n_cells_pre_doublet_filter <- ncol(combined_filtered)

# Subset to keep only cells where scDblFinder_class is "singlet"
combined_singlets <- subset(combined_filtered, subset = scDblFinder_class == "singlet")

n_doublets_removed <- n_cells_pre_doublet_filter - ncol(combined_singlets)
pct_doublets_removed <- round(n_doublets_removed / n_cells_pre_doublet_filter * 100, 2)

print(paste("Cells before doublet filtering:", n_cells_pre_doublet_filter))
print(paste("Doublets removed:", n_doublets_removed, paste0("(", pct_doublets_removed, "%)")))
print(paste("Cells remaining (Singlets):", ncol(combined_singlets)))
```

# VISUALIZATION AND SUMMARY
Plots are generated using the pre-filtered object (combined_filtered) to show where the doublets were located before removal.

```{r}
# Summarize doublets per sample (using the pre-filtered object for reporting)
doublet_summary <- combined_filtered@meta.data %>%
 group_by(Sample_ID) %>%
 summarise(
  total_cells_pre_filter = n(),
  n_doublets = sum(scDblFinder_class == "doublet"),
  pct_doublets = round(100 * n_doublets / total_cells_pre_filter, 2)
 ) %>%
 arrange(Sample_ID)

doublet_summary
```

```{r}
# UMAP plot highlighting doublets (using the pre-filtered object for plotting)
p_umap_doublet <- DimPlot(
 combined_filtered,
 group.by = "scDblFinder_class",
 reduction = "umap",
 pt.size = 0.5,
 raster = FALSE,
 cols = c("singlet" = "grey70", "doublet" = "red") 
) +
 labs(title = "UMAP Highlighting scDblFinder Doublets (Pre-Filter)") +
 theme_minimal(base_size = 12) +
  guides(color = guide_legend(override.aes = list(size = 3))) 

print(p_umap_doublet)
ggsave("plots/03_UMAP_doublets_scDblFinder.png", plot = p_umap_doublet, width = 7, height = 6, dpi = 300)
```

# SAVE THE FILTERED SINGLETS OBJECT

```{r}
# Save the new object containing ONLY singlets
saveRDS(combined_singlets, file = SAVE_OBJECT_PATH)
print("Step 3 (Doublet Filtering) Complete. Object saved containing only singlets.")
```
