---
title: "R Notebook - Quality Check and Visualization"
author : Srivalli Kolla
output: html_notebook
---

# SETUP AND LOAD LIBRARIES

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tibble)
library(reshape2)
library(inflection)
library(DropletUtils)
```

# DEFINE DATA PATHS AND METADATA

```{r}
# List of sample names 
sample_names <- c(
 "Sample1_Control", "Sample2_Control", "Sample3_Treated", "Sample4_Treated"
)

# List of full file paths to the Cell Ranger filtered_feature_bc_matrix directory for each sample
data_paths <- paste0("path/to/your/data/location/", sample_names, "/filtered_feature_bc_matrix/") # update with your path

# Define the custom metadata table - Update values to match your experimental design 
meta_info <- data.frame(
 sample_id_short = paste0("S", 1:length(sample_names)),
 Sample_ID = sample_names,
 Sex = rep(c("male", "female"), length(sample_names)/2),
 Genotype = rep(c("WildType", "Mutant"), each = length(sample_names)/2),
 Condition = rep(c("Control", "Treated"), each = length(sample_names)/2),
 stringsAsFactors = FALSE 
)
```

# IMPORT DATA AND CALCULATE QC METRICS PER SAMPLE

```{r}
seurat_list <- list()
# Create a loop to go through all the samples
for (i in seq_along(sample_names)) {
 sample_tag <- sample_names[i]
 data <- Read10X(data.dir = data_paths[i])

 # Make barcodes unique by adding a suffix (e.g., _Sample1_Control)
 colnames(data) <- paste0(colnames(data), "_", sample_tag)

 seu <- CreateSeuratObject(counts = data, project = sample_tag)
 seu$Sample_ID <- sample_tag # Store full sample name
 seu$sample_short <- meta_info$sample_id_short[i] # Store short sample ID

    # NOTE: Adjust the pattern for mitochondrial genes based on your species: Human: "^MT-" | Mouse: "^mt-" | Ribosomal: "^Rps|^Rpl"
 
 seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^mt-")
 seu[["percent.ribo"]] <- PercentageFeatureSet(seu, pattern = "^Rps|^Rpl")

 # Add metadata from meta_info
 meta_row <- meta_info[meta_info$Sample_ID == sample_tag, ]

 if (nrow(meta_row) == 1) {
  for (col in colnames(meta_row)) {
   seu[[col]] <- meta_row[[col]][1] # Use [1] to handle the data frame structure
  }
 } else {
  warning(paste("Metadata not found for:", sample_tag))
 }

 seurat_list[[sample_tag]] <- seu
}
```

# MERGE RAW OBJECTS FOR COMBINED QC VISUALIZATION

```{r}
# Merge all the samples information
combined_raw <- merge(seurat_list[[1]], y = seurat_list[-1])
combined_raw

# For large datasets, use the following line to clear memory *after* running the merge
# rm(seurat_list)
```

# VISUALIZATION OF QC METRICS (TO DETERMINE FILTERING THRESHOLDS)

```{r}
# PLOT A: Violin Plots by Sample
p_violin_before <- VlnPlot(
 combined_raw,
 features = c("nFeature_RNA", "nCount_RNA", "percent.mt","percent.ribo"),
 group.by = "Sample_ID", pt.size = 0, raster=FALSE, ncol = 4
)
print(p_violin_before)
ggsave("plots/01_QC_ViolinPlots_Raw.png", p_violin_before, width = 14, height = 6)
```

```{r}
# PLOT B: Bar Plot of Raw Cell Counts
cell_counts_before <- as.data.frame(table(combined_raw$Sample_ID))
colnames(cell_counts_before) <- c("Sample_ID", "CellCount")

bar_1 <- ggplot(cell_counts_before, aes(x = Sample_ID, y = CellCount)) +
 geom_bar(stat = "identity") +
 geom_text(aes(label = CellCount), vjust = -0.5, size = 3) +
 theme_minimal() +
 labs(title = "Raw Cell Counts Before Filtering", x = "Sample", y = "Number of Cells") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(bar_1)
ggsave("plots/01_CellCounts_Raw.png", bar_1, width = 8, height = 6)
```

```{r}
# PLOT C: Barcode Rank Plot
# Extract raw counts 
raw_counts <- combined_raw[["RNA"]]@counts

# Calculate barcode ranks (uses the EDE internally for knee/inflection)
br.out <- barcodeRanks(raw_counts)

# Extract points for annotation
knee_point <- metadata(br.out)$knee
inflection_point <- metadata(br.out)$inflection

# Create plot data
df_elbow <- data.frame(
    rank = br.out$rank, 
    counts = br.out$total
)

p_knee <- ggplot(df_elbow, aes(x = rank, y = counts)) +
 geom_line(color = "purple", size = 1) +
 scale_y_log10() +
 scale_x_log10() + # Log-log scale for clear knee visualization
 
 # Highlight the automatically calculated Knee point
 geom_hline(yintercept = knee_point, linetype = "dashed", color = "red", linewidth = 1) +
 annotate("text", x = 10, y = knee_point, 
           label = paste0("Knee (UMI Min): ", formatC(knee_point, format="d", big.mark=",")), 
           color = "red", size = 4, hjust = 0, vjust = -0.5) +

 labs(title = "Canonical Barcode Rank Plot (UMI Counts)", 
       subtitle = "Red line is the estimated knee point (minimum UMI cutoff).",
       x = "Cell Rank (Log Scale)", 
       y = "Total Counts (nCount_RNA) (Log Scale)") +
 theme_minimal(base_size = 13)

print(p_knee)
ggsave("plots/01_QC_BarcodeRank_KneePlot.png", p_knee, width = 8, height = 6, dpi = 300)
```

```{r}
# Function to calculate and plot EDE inflection point on a QC metric
plot_qc_inflection_ede <- function(combined_raw, feature_name, bin_width_val, plot_title, reverse_sort = FALSE) {
    
    # Prepare data vector (sort is crucial for inflection package)
    qc_vector <- combined_raw@meta.data[[feature_name]]
    qc_vector_sorted <- sort(qc_vector, decreasing = reverse_sort)
    
    # Calculate inflection points using EDE
    inflect_res <- tryCatch({
        # Use EDE (index 2) calculation
        inflection::findiplist(seq_along(qc_vector_sorted), qc_vector_sorted, index = TRUE)
    }, error = function(e) {
        warning(paste("Inflection calculation failed for", feature_name, ":", e$message))
        return(NULL)
    })
    
    if (is.null(inflect_res) || nrow(inflect_res) < 2) {
        message(paste("Skipping EDE plot for", feature_name, "due to calculation failure or insufficient results."))
        return(NULL)
    }
    
    # Extract EDE value (typically row 2 for EDE)
    inflection_index_EDE_round <- round(inflect_res[2, 1])
    inflection_value_EDE <- qc_vector_sorted[inflection_index_EDE_round]

    # Create histogram plot
    df <- data.frame(qc_value = qc_vector)
    
    # Determine max height for annotation positioning
    h_counts <- table(cut(df$qc_value, breaks = seq(min(df$qc_value, na.rm=TRUE), max(df$qc_value, na.rm=TRUE), by = bin_width_val)))
    max_y <- max(h_counts, na.rm=TRUE) * 0.9
    
    p <- ggplot(df, aes(x = qc_value)) +
        geom_histogram(binwidth = bin_width_val, fill = "grey80", color = "black") +
        
        # Highlight the suggested EDE cutoff
        geom_vline(xintercept = inflection_value_EDE, color = "blue", linetype = "dashed", linewidth = 1) +
        
        labs(title = plot_title,
             subtitle = paste0("Suggested EDE Cutoff: ", round(inflection_value_EDE, ifelse(inflection_value_EDE > 10, 0, 2))),
             x = feature_name,
             y = "Number of Nuclei") +
        
        annotate("text", x = inflection_value_EDE, y = max_y, 
                 label = paste0("EDE: ", round(inflection_value_EDE, ifelse(inflection_value_EDE > 10, 0, 2))), 
                 color = "blue", angle = 90, vjust = -0.5) +
        
        theme_minimal(base_size = 13)
    
    print(p)
    
    # Save the plot
    ggsave(paste0("plots/01_QC_Inflection_", feature_name, ".png"), plot = p, width = 8, height = 6, dpi = 300)
    
    # Return the suggested value
    return(inflection_value_EDE)
}
```

```{r}
# PLOT D : QC Metrics Histograms with EDE Inflection

# 1. Total Counts (nCount_RNA) - Use EDE on histogram for comparison with knee point
p_counts <- plot_qc_inflection_ede(combined_raw, "nCount_RNA", bin_width_val = 500, plot_title = "Total Counts (nCount_RNA)", reverse_sort = FALSE)

# 2. Number of Features (nFeature_RNA) - Lower cutoff suggestion
p_features <- plot_qc_inflection_ede(combined_raw, "nFeature_RNA", bin_width_val = 200, plot_title = "Detected Genes (nFeature_RNA)", reverse_sort = FALSE)

# 3. Mitochondrial percentage (percent.mt) - Upper cutoff suggestion
# NOTE: bin_width = 1 is standard for percentages.
p_mt <- plot_qc_inflection_ede(combined_raw, "percent.mt", bin_width_val = 1, plot_title = "Mitochondrial Percentage (%MT)", reverse_sort = FALSE)

# 4. Ribosomal percentage (percent.ribo) - Upper cutoff suggestion
p_ribo <- plot_qc_inflection_ede(combined_raw, "percent.ribo", bin_width_val = 1, plot_title = "Ribosomal Percentage (%Ribo)", reverse_sort = FALSE)


# C. Combine all EDE Histogram plots for easy comparison
combined_ede_plots <- (p_counts / p_features) | (p_mt / p_ribo)
ggsave("plots/01_QC_EDE_Histograms_Combined.png", combined_ede_plots, width = 16, height = 12, dpi = 300)
```

The Extremum Distance Estimator (EDE) line is a data-driven tool that finds the point of maximum curvature on the distribution of your QC metrics. This value provides a reproducible threshold suggesting the boundary between high-quality profiles and debris/stressed cells.

 EDE Guidance Summary:
 - For MINIMUM cutoffs (nCount_RNA, nFeature_RNA), EDE suggests the point separating background noise/debris where expression profile becomes too sparse, indicating poor-quality or fragmented nuclei.
 - For MAXIMUM cutoffs (percent.mt, percent.ribo), EDE suggests the point where the high-quality distribution starts its sharp decline indicating compromised or dying cells beyond this point.

# SAVE THE RAW OBJECT

```{r}
saveRDS(combined_raw, file = "data/processed/combined_raw_unfiltered.rds") 
print("Step 1 (QC and Visualization) Complete. Raw Seurat object saved to 'data/processed/'.")
```

