---
title: "R Notebook - Filtering and Normalization"
author: Srivalli Kolla
output: html_notebook
---

# SETUP AND LOAD LIBRARIES

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# LOAD RAW, UNFILTERED DATA

```{r}
INPUT_OBJECT_PATH <- "data/processed/combined_raw_unfiltered.rds"
SAVE_OBJECT_PATH <- "data/processed/combined_filtered_normalized.rds"

if (!file.exists(INPUT_OBJECT_PATH)) {
    stop("ERROR: Raw unfiltered object not found. Run 01_qc_visualization.Rmd first!")
}
combined_raw <- readRDS(INPUT_OBJECT_PATH)

print(paste("Loaded raw object with", ncol(combined_raw), "nuclei and", nrow(combined_raw), "features."))
```

# DEFINE AND APPLY FILTERING THRESHOLDS

```{r}
# User-Defined Thresholds (EDIT THESE VALUES - Based on EDE calculated in 01_qc_visualization.Rmd)
min_nFeature_RNA <- 300   # Minimum genes detected
max_nFeature_RNA <- 7000  # Maximum genes detected 
max_percent_mt <- 20      # Maximum mitochondrial percentage allowed
max_percent_ribo <- 45    # Maximum ribosomal percentage allowed 

# Apply Filters
combined_filtered <- subset(combined_raw, subset = 
    nFeature_RNA > min_nFeature_RNA & 
    nFeature_RNA < max_nFeature_RNA & 
    percent.mt < max_percent_mt &
    percent.ribo < max_percent_ribo
)

print(paste("Raw cells:", ncol(combined_raw)))
print(paste("Filtered cells:", ncol(combined_filtered)))

# Calculate percentage of cells retained
retained_pct <- round(ncol(combined_filtered) / ncol(combined_raw) * 100, 2)
print(paste("Percentage of cells retained:", retained_pct, "%"))
```

# QC VISUALIZATION AFTER FILTERING

```{r}
# PLOT : Violin Plots After Filtering
p_violin_after <- VlnPlot(
  combined_filtered,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt","percent.ribo"),
  group.by = "Sample_ID", pt.size = 0.1, raster=FALSE, ncol = 4
) + labs(title = "QC Metrics After Filtering")
print(p_violin_after)
ggsave("plots/02_QC_ViolinPlots_Filtered.png", p_violin_after, width = 14, height = 6)
```

# NORMALIZATION AND VARIANCE STABILIZATION (SCTransform)

SCTransform performs:
  1. Normalization (using ambient noise modeling)
  2. Variance stabilization
  3. Scaling (automatically stores counts in the 'SCT' assay)

We are not regressing out cell cycle (using vars.to.regress). We are only saving the scores for later inspection.
```{r}
# Set raw counts back as default before running SCTransform
DefaultAssay(combined_filtered) <- "RNA"

# SCTransform performs filtering, normalization, and scaling.
combined_filtered <- SCTransform(combined_filtered,
                vars.to.regress = c("percent.mt"), 
                verbose = FALSE)

# Note: The 'percent.mt' variable is regressed out. Cell cycle effects are retained.
```

# ADD CELL CYCLE SCORING AND CHECK

Score cells for S and G2M phases and check if cell cycle phase is highly correlated with low QC metrics.

```{r}
# Mouse cell cycle gene sets
s.genes.mouse <- c("Mcm5","Pcna","Tyms","Fen1","Mcm2","Mcm4","Rrm1","Ung","Gins2","Mcm6","Cdca7","Dtl","Prim1","Uhrf1","Mcm3","Mcm7","Slbp","Cdc45","Tk1","Lig1","Rpa2","Cdc6","Exo1","Tipin","Dscc1","Blm","Casp8ap2","Usp1","Clspn","Polr1d","Hells","Rfc2","Rpa3","Rrm2","Nasp","Rad51ap1","Gmnn","Wdr76","Slk","Cdc7","Dck")
g2m.genes.mouse <- c("Hmgb2","Cdc20","Bub1","Cdk1","Tpx2","Nusap1","Ckap2l","Ckap2","Aurkb","Birc5","Kif11","Anp32e","Tubb4b","Gtse1","Kif20b","Hjurp","Cdca3","Cdc25c","Ckap5","Cdc20","Esco2","Hmgb2","Cenpf","Ttk","Kif2c","Ndc80","Cdc45","Tacc3","Cdc6","Pimreg","Kif4a","Nuf2","Cenpe","Cdca8","Ect2","Kif23","Cdc20","Cks2","Nek2","Cdca2","Cks1b","Cdc25a")

# Cell cycle scoring: uses the currently active assay (SCT) by default
combined_filtered <- CellCycleScoring(
combined_filtered,
s.features = s.genes.mouse,
g2m.features = g2m.genes.mouse,
set.ident = FALSE # Do not change active identity
)

print("Cell cycle scores (S.Score, G2M.Score, Phase) have been added to metadata.")

# VISUALIZATION CHECK 
p_cc_vln <- VlnPlot(combined_filtered,
       features = c("S.Score", "G2M.Score"),
       group.by = "Phase", # Check scores by the assigned phase
       pt.size = 0, ncol = 2)+
             labs(title = "S and G2M Scores by Assigned Phase")

print(p_cc_vln)
ggsave("plots/02_VlnPlot_CC_Scores.png", p_cc_vln, width = 8, height = 4)
```

# SAVE THE FILTERED AND NORMALIZED OBJECT

```{r}
saveRDS(combined_filtered, file = SAVE_OBJECT_PATH)
print("Step 2 (Filtering and Normalization) Complete. Object saved with Cell Cycle scores.")
```

