---
title: "R Notebook - Filtering and Normalization"
author: Srivalli Kolla
output: html_notebook
---

# SETUP AND LOAD LIBRARIES

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# LOAD RAW, UNFILTERED DATA

```{r}
INPUT_OBJECT_PATH <- "data/processed/combined_raw_unfiltered.rds"
SAVE_OBJECT_PATH <- "data/processed/combined_filtered_normalized.rds"

if (!file.exists(INPUT_OBJECT_PATH)) {
    stop("ERROR: Raw unfiltered object not found. Run 01_qc_visualization.Rmd first!")
}
combined_raw <- readRDS(INPUT_OBJECT_PATH)

print(paste("Loaded raw object with", ncol(combined_raw), "nuclei and", nrow(combined_raw), "features."))
```

# DEFINE AND APPLY FILTERING THRESHOLDS

```{r}
# User-Defined Thresholds (EDIT THESE VALUES - Based on EDE calculated in 01_qc_visualization.Rmd)
min_nFeature_RNA <- 300   # Minimum genes detected
max_nFeature_RNA <- 7000  # Maximum genes detected 
max_percent_mt <- 20      # Maximum mitochondrial percentage allowed
max_percent_ribo <- 45    # Maximum ribosomal percentage allowed 

# Apply Filters
combined_filtered <- subset(combined_raw, subset = 
    nFeature_RNA > min_nFeature_RNA & 
    nFeature_RNA < max_nFeature_RNA & 
    percent.mt < max_percent_mt &
    percent.ribo < max_percent_ribo
)

print(paste("Raw cells:", ncol(combined_raw)))
print(paste("Filtered cells:", ncol(combined_filtered)))

# Calculate percentage of cells retained
retained_pct <- round(ncol(combined_filtered) / ncol(combined_raw) * 100, 2)
print(paste("Percentage of cells retained:", retained_pct, "%"))
```

# QC VISUALIZATION AFTER FILTERING

```{r}
# PLOT : Violin Plots After Filtering
p_violin_after <- VlnPlot(
  combined_filtered,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt","percent.ribo"),
  group.by = "Sample_ID", pt.size = 0.1, raster=FALSE, ncol = 4
) + labs(title = "QC Metrics After Filtering")
print(p_violin_after)
ggsave("plots/02_QC_ViolinPlots_Filtered.png", p_violin_after, width = 14, height = 6)
```

# NORMALIZATION AND VARIANCE STABILIZATION (SCTransform)

SCTransform performs:
  1. Normalization (using ambient noise modeling)
  2. Variance stabilization
  3. Scaling (automatically stores counts in the 'SCT' assay)

```{r}
# # SCTransform
combined_filtered <- SCTransform(combined_filtered, 
                                 vars.to.regress = c("percent.mt"), 
                                 verbose = FALSE)

# Note: The 'percent.mt' variable is regressed out to remove variance driven by cell quality differences.
```

# SAVE THE FILTERED AND NORMALIZED OBJECT

```{r}
saveRDS(combined_filtered, file = SAVE_OBJECT_PATH)
print("Step 2 (Filtering and Normalization) Complete. Object saved to 'data/processed/'.")
```

