---
title: "Compositional Analysis - Cell Proportion & Differential Abundance (MiloR)"
author: Srivalli Kolla
output: html_notebook
---

#  Compositional Analysis and Differential Abundance

This script performs two types of compositional analysis:
1. **Cell Type Proportion Analysis:** Standard analysis using Wilcoxon tests on cell proportions per sample.
2. **Differential Abundance (DA) Analysis:** Advanced, local analysis using the **MiloR** package to identify changes in cell density (DA neighborhoods) between genotypes.

## 1. Setup and Data Loading

```{r}
library(miloR)
library(SingleCellExperiment)
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggpubr) 
library(viridis)
library(tidyr) 
library(ggrepel)
library(rstatix)
```

# Load Data

```{r}
# Define relative output paths
data_out_dir <- "data/compositional_analysis"
plots_out_dir <- "plots/compositional_analysis"
dir.create(data_out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plots_out_dir, showWarnings = FALSE, recursive = TRUE)

singlets <- readRDS("path/to/combined_de_annotated_MAST.rds")
cat("Seurat object loaded with", ncol(singlets), "nuclei.\n")

DefaultAssay(singlets) <- "SCT"

# Convert to SingleCellExperiment and set required columns
sce <- as.SingleCellExperiment(singlets, assay = "RNA")
colData(sce)$Sample <- colData(sce)$Sample_ID
colData(sce)$Genotype <- colData(sce)$Genotype
```

# Global Cell Proportion Analysis

## Calculate Proportions and Test
```{r}
# Count cells and calculate proportions using DE_Annotation_broad
prop_df <- as.data.frame(table(sce$Sample_ID, sce$DE_Annotation_broad, sce$Genotype)) %>%
  setNames(c("Sample_ID", "CellType", "Genotype", "Count")) %>%
  group_by(Sample_ID) %>%
  mutate(Prop = Count / sum(Count)) %>%
  ungroup()

# Wilcoxon tests per broad cell type
results_prop <- prop_df %>%
  group_by(CellType) %>%
  wilcox_test(Prop ~ Genotype) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  rename(FDR = p.adj) %>%
  select(CellType, p, FDR, p.signif) %>%
  arrange(FDR)

# Save stats
write_csv(results_prop, file.path(data_out_dir, "celltype_proportion_stats_broad.csv"))
results_prop
```


## Visualization: Proportion Boxplots
```{r}
p_prop <- ggplot(prop_df, aes(x = Genotype, y = Prop, fill = Genotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 14) +
  labs(y = "Proportion of Nuclei per Sample", title = "Global Cell Type Proportions (Broad)") +
  theme(strip.text = element_text(face = "bold"), legend.position = "none") +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = 0.98, size = 4)

ggsave(file.path(plots_out_dir, "celltype_proportion_boxplots_broad.png"), plot = p_prop, width = 10, height = 8, dpi = 300)
p_prop
```

# Differential Abundance Analysis (MiloR)

## Build Milo Object, Graph, and Neighborhoods
```{r}
milo <- Milo(sce)
milo <- buildGraph(milo, k = 20, d = 30, reduced.dim = "PCA") 
milo <- makeNhoods(milo, prop = 0.03, k = 20, d = 30, refined = TRUE)
milo <- countCells(milo, meta.data = as.data.frame(colData(sce)), sample = "Sample_ID")
```

## Differential Abundance Testing and Mapping
```{r}
# 1. Prepare sample metadata
sample_meta <- distinct(as.data.frame(colData(sce)[, c("Sample_ID", "Genotype")]))
rownames(sample_meta) <- sample_meta$Sample_ID

# 2. Run DA test: Define model and contrast
design <- ~ Genotype
model_contrasts <- "Genotype1" 
da_results <- testNhoods(milo, design.df = sample_meta, design = design, model.contrasts = model_contrasts)

# 3. Map neighborhoods to majority broad cell type
nhood_mat <- as.matrix(milo@nhoods)
common_cells <- intersect(rownames(nhood_mat), colnames(sce))
nhood_mat <- nhood_mat[common_cells, , drop = FALSE]

get_majority_celltype <- function(nhood_index, annotation_col) {
  nh_cells <- common_cells[which(nhood_mat[, nhood_index] > 0)]
  if (length(nh_cells) == 0) return(NA)
  names(sort(table(colData(sce)[nh_cells, annotation_col]), decreasing = TRUE))[1]
}

da_results <- da_results %>%
  mutate(
    CellType_broad = sapply(seq_len(ncol(nhood_mat)), get_majority_celltype, annotation_col = "DE_Annotation_broad")
  )
```

## Summarize and Save Results
```{r}
# Function to create structured summary tables
summarize_da <- function(df, celltype_col) {
  df %>%
    filter(!is.na(.data[[celltype_col]])) %>%
    group_by(.data[[celltype_col]]) %>%
    summarize(
      mean_logFC = mean(logFC, na.rm = TRUE),
      n_total_nhoods = n(),
      n_signif_nhoods = sum(PValue < 0.05, na.rm = TRUE),
      min_pval = min(PValue, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rename(CellType = all_of(celltype_col))
}

da_summary_broad <- summarize_da(da_results, "CellType_broad")

# Save DA results and summary
write_csv(da_results, file.path(data_out_dir, "DA_miloR_results_broad.csv"))
write_csv(da_summary_broad, file.path(data_out_dir, "DA_summary_broad.csv"))

cat("\nBroad Summary:\n")
da_summary_broad
```

# Visualization

## Volcano Plot of Neighborhoods
```{r}
# Plot function for volcano
volcano_plot <- function(df, celltype_col, title, out_file, p_cutoff = 0.05) {
  df <- df %>%
    mutate(Significant = ifelse(PValue < p_cutoff, "Yes", "No"))
  df_labels <- df %>% filter(Significant == "Yes") %>% group_by(.data[[celltype_col]]) %>% slice_max(order_by = -log10(PValue), n = 1) %>% ungroup()
  
  p <- ggplot(df, aes(x = logFC, y = -log10(PValue), color = .data[[celltype_col]], shape = Significant)) +
    geom_point(alpha = 0.7, size = 3) +
    geom_text_repel(data = df_labels, aes(label = .data[[celltype_col]]), size = 3, color = "black", max.overlaps = 5) +
    scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
    geom_hline(yintercept = -log10(p_cutoff), linetype = "dashed", color = "grey50") +
    theme_minimal(base_size = 14) +
    labs(title = title, x = "logFC (Genotype vs Control)", y = expression(-log[10](P-value)), color = "Cell Type")
  
  ggsave(out_file, plot = p, width = 10, height = 7, dpi = 300)
}

volcano_plot(da_results, "CellType_broad", "DA neighborhoods (Broad Annotation)", file.path(plots_out_dir, "DA_volcano_broad.png"))
```

## Bar Plot: Count of Significant Neighborhoods

```{r}
# Plot function for significant bar plot
barplot_da_sig <- function(df, celltype_col, title, out_file) {
  summary_df <- df %>%
    filter(!is.na(.data[[celltype_col]])) %>%
    group_by(.data[[celltype_col]]) %>%
    summarize(
      n_signif = sum(PValue < 0.05),
      mean_logFC_sig = mean(logFC[PValue < 0.05], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(mean_logFC_sig = ifelse(is.na(mean_logFC_sig), 0, mean_logFC_sig))
  
  p <- ggplot(summary_df, aes(x = reorder(.data[[celltype_col]], n_signif), y = n_signif, fill = mean_logFC_sig)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Mean logFC (signif only)") +
    labs(title = title, y = "Number of Significant Neighborhoods", x = NULL) +
    theme_minimal(base_size = 13)
  
  ggsave(out_file, plot = p, width = 8, height = 5, dpi = 300)
}

barplot_da_sig(da_results, "CellType_broad", "Significant Neighborhoods per Broad Cell Type", file.path(plots_out_dir, "DA_bar_broad_sig.png"))
```

## Faceted Boxplot of logFC

```{r}
# Plot function for faceted boxplot
facet_boxplot_da <- function(df, celltype_col, title, out_file) {
  summary_df <- df %>%
    filter(!is.na(.data[[celltype_col]])) %>%
    group_by(.data[[celltype_col]]) %>%
    summarize(mean_logFC = mean(logFC, na.rm = TRUE), max_logFC = max(logFC, na.rm = TRUE), min_pval = min(PValue, na.rm = TRUE),
              p.signif = case_when(min_pval <= 0.001 ~ "***", min_pval <= 0.01  ~ "**", min_pval <= 0.05  ~ "*", TRUE ~ ""),
              .groups = "drop") %>%
    mutate(label_text = ifelse(p.signif != "", paste0(p.signif, "\n(P=", signif(min_pval, 3), ")"), ""))
  
  df[[celltype_col]] <- factor(df[[celltype_col]], levels = summary_df %>% arrange(desc(mean_logFC)) %>% pull(!!sym(celltype_col)))
  
  p <- ggplot(df, aes(x = "Comparison", y = logFC, fill = logFC)) +
    geom_boxplot(outlier.shape = 1, alpha = 0.7) +
    facet_wrap(as.formula(paste("~", celltype_col)), scales = "free_y") +
    geom_text(data = summary_df, aes(x = 1, y = max_logFC + 0.1, label = label_text), color = "black", size = 4, lineheight = 0.9, inherit.aes = FALSE) +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name = "logFC") +
    labs(title = title, x = NULL, y = "Neighborhood logFC (Genotype vs Control)") +
    theme_minimal(base_size = 14) +
    theme(strip.text = element_text(face = "bold"), plot.title = element_text(face = "bold", hjust = 0.5), axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom")
  
  ggsave(out_file, plot = p, width = 12, height = 8, dpi = 300)
}

facet_boxplot_da(da_results, "CellType_broad", "Faceted DA logFC per Broad Cell Type", file.path(plots_out_dir, "DA_boxplot_broad_facet.png"))
```

